Sequence Labeling for Gait Analysis using LSTM

### 2015-05-09 paper/paper.tex
Fix typos

\begin{abstract}
    In this paper, we propose a new data-based approach to gait event detection 
    based on a Long Short-Term Memory Recurrent Neural Network.
    Our approach decisively outperforms existing methods such as the Foot 
    Velocity Algorithm and feed-forward Neural Networks with sliding window 
    in terms of the number of misclassified events and has a comparable 
    performance in terms of bias and variance for correctly classified events.
    Further advantages of our approach are that it does not rely on
    questionable heuristics and that it does not require heavy preprocessing 
    of the data.
    Given the promising initial results, we believe that it should be possible 
    to further improve the current state of the art for data-based event 
    detection using a Long Short-Term Memory network.
\end{abstract}

@@

There are three basic approaches to event detection.
The first approach uses visual inspection to manually label the events.
Although quite accurate, the cost associated with this method is prohibitive for
all but the smallest amounts of data.
For this reason, it is not usually used as a stand-alone method but rather as a
postprocessing step for automated event detection systems or as a means to
generate small sets of hand-labeled test data.
The second approach uses dedicated hardware such as force plates that measure
ground reaction forces and foot switches that are pressed when the foot is in
contact with the ground.
Due to its high accuracy, hardware-based methods are considered to be
state of the art for gait event identification.
However, their usefulness is limited by the fact that many laboratories do not
have access to the necessary equipment.
Furthermore, there is a risk of affecting the gait because some of the devices
require the modification of normal footware.
The third approach consists in automatated event detection based on solely on
the data.
If succesful, this approach is superior to the other two because it scales
easily, does not require additional equipment and does not pose a risk of
affecting the gait.

Given their apparent advantages, it is not surprising that several data-based
methods have been proposed in the literature.
Although some of those methods achieve results that are accurate enough to be
useful in practice, they all have drawbacks such as relying heavily on
questionable heuristics or requiring an undue amount of data preprocessing.
For this reason, we present a new approach to gait event detection using a
Long Short-Term Memory (LSTM) recurrent neural network (RNN).
We believe that our method is superior to existing approaches both in terms of
accuracy and in terms of only requirying a small amount of training data and
preprocessing.

@@

This paper is organized as follows:
Section~\ref{sec:Previous Work}
discusses some of the existing data-based methods for event detection;
Section~\ref{sec:LSTM}
gives a short overview over LSTM networks in general;
Section~\ref{sec:Data}
contains a description the dataset;
Sections~\ref{sec:Network Architecture} and \ref{sec:Network Training}
describe the architecture and training of our network;
Section~\ref{sec:Results}
presents the experimental results and compares them to existing baselines;
Section~\ref{sec:Conclusion and Future Work}
concludes and shows possible paths for future work.

@@

\subsection{Parameter search}
Initially we wanted to determine what values to use for the training of our network to achieve the best possible performance. In order to do this, we trained our network for a few epochs (10) with a  number of different parameters to see its performance.
For each parameter set, we ran the training three times, in order to average among the realizations and obtain results less susceptible to the initial conditions.
All results from this section used the training half of Dataset 1, and training was implemented as stochastic gradient descent (SGD), with each datapoint corresponding to a sequence length of 500 samples (5 seconds).

\begin{figure}[H]
 \centering
 \includegraphics[scale=.5]{./figures/parameterSearch.png}
 % parameterSearch.png: 0x0 pixel, 300dpi, 0.00x0.00 cm, bb=
 \caption{Convergence of the network in the first 10 epochs as a function of different parameters.}
 \label{fig:paramSearch}
\end{figure}

@@

\begin{figure}[H]
 \centering
 \includegraphics[scale=.4]{./figures/Test2_2Train2_1Compare.png}
 % Test1_1Train1_2Compare.png: 0x0 pixel, 300dpi, 0.00x0.00 cm, bb=
 \caption{Comparison of Miller vs. LSTM network for various training intervals, for dataset 2.}
 \label{fig:halfAndHalfUsers}
\end{figure}

>>>>

\begin{abstract}
    In this paper, we propose a new data-based approach to gait event detection 
    using a Long Short-Term Memory Recurrent Neural Network.
    Our approach decisively outperforms existing methods such as the Foot 
    Velocity Algorithm and feed-forward Neural Networks with sliding window 
    in terms of the number of misclassified events and has a comparable 
    performance in terms of bias and variance for correctly classified events.
    Further advantages of our approach are that it does not rely on
    questionable heuristics and that it does not require heavy preprocessing 
    of the data.
    Given the promising initial results, we believe that it should be possible 
    to even further improve the current state of the art for data-based event 
    detection using a Long Short-Term Memory network.
\end{abstract}

@@

There are three basic approaches to event detection.
The first approach uses visual inspection to manually label the events.
Although quite accurate, the cost associated with this method is prohibitive for
all but the smallest amounts of data.
For this reason, it is not usually used as a stand-alone method but rather as a
postprocessing step for automated event detection systems or as a means to
generate small amounts of hand-labeled test data.
The second approach uses dedicated hardware such as force plates that measure
ground reaction forces and foot switches that are pressed when the foot is in
contact with the ground.
Due to their high accuracy, hardware-based methods are considered to be
state of the art for gait event identification.
However, their usefulness is limited by the fact that many laboratories do not
have access to the necessary equipment.
Furthermore, there is a risk of affecting the gait because some of the devices
require the modification of normal footware.
The third approach consists in automated event detection based solely on
the data.
If successful, this approach is superior to the other two because it scales
easily, does not require additional equipment and does not pose a risk of
affecting the gait.

Given these apparent advantages, it is not surprising that several data-based
methods have been proposed in the literature.
Although some of those methods achieve results that are accurate enough to be
useful in practice, they all have drawbacks such as relying heavily on
questionable heuristics or requiring an undue amount of data preprocessing.
For this reason, we present a new approach to gait event detection using a
Long Short-Term Memory (LSTM) recurrent neural network (RNN).
We believe that our method is superior to existing approaches both in terms of
accuracy and in terms of only requirying a small amount of training data and
preprocessing.

@@

This paper is organized as follows:
Section~\ref{sec:Previous Work}
discusses some of the existing data-based methods for event detection;
Section~\ref{sec:LSTM}
gives a short overview over LSTM networks in general;
Section~\ref{sec:Data}
contains a description the dataset;
Section~\ref{sec:Computational Resources}
shows states what computational resources were used for this project;
Sections~\ref{sec:Network Architecture} and \ref{sec:Network Training}
describe the architecture and training of our network;
Section~\ref{sec:Results}
presents the experimental results and compares them to existing baselines;
Section~\ref{sec:Conclusion and Future Work}
concludes and shows possible paths for future work.

@@

% \subsection{Parameter search}
Initially we wanted to determine what values to use for the training of our network to achieve the best possible performance. In order to do this, we trained our network for a few epochs (10) with a  number of different parameters to see its performance.
For each parameter set, we ran the training three times, in order to average among the realizations and obtain results less susceptible to the initial conditions.
All results from this section used the training half of Dataset 1, and training was implemented as stochastic gradient descent (SGD), with each datapoint corresponding to a sequence length of 500 samples (5 seconds).

\begin{figure}[H]
 \centering
 \includegraphics[scale=.45]{./figures/parameterSearch.png}
 % parameterSearch.png: 0x0 pixel, 300dpi, 0.00x0.00 cm, bb=
 \caption{Convergence of the network in the first 10 epochs as a function of different parameters.}
 \label{fig:paramSearch}
\end{figure}

@@

\begin{figure}[H]
 \centering
 \includegraphics[scale=.45]{./figures/Test2_2Train2_1Compare.png}
 % Test1_1Train1_2Compare.png: 0x0 pixel, 300dpi, 0.00x0.00 cm, bb=
 \caption{Comparison of Miller vs. LSTM network for various training intervals, for dataset 2.}
 \label{fig:halfAndHalfUsers}
\end{figure}
